D3064R0: How to Avoid OOTA Without Really Trying

### 2024-03-29 ootavacuous.tex
Additional minor edits
Here are a few very minor last-minute changes.  The biggest one is
probably my decision that "many-to-one" ought to be hyphenated because
it is not a modifier-noun combination.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

This section uncovers some semantic-dependency complexities.
Section~\ref{sec:Semantic Dependencies and Source Code}
shows that semantic dependencies are functions of executions,
rather than being strict functions of the source code.
Section~\ref{sec:Semantic Dependencies Can Be Many To One}
shows that semantic dependencies do not necessarily extend
from a single load to a single store, but can instead involve
multiple loads.
Section~\ref{sec:Semantic Dependencies Affected by Cross-Thread Optimizations}
shows that semantic dependencies can be affected by cross-thread
optimizations.
Sections~\ref{sec:Semantic Dependencies Affected by if Statements}
and~\ref{sec:Semantic Dependencies Not Affected by if Statements}
show that \co{if} statements can have surprising effects on semantic
dependencies, up to and including eliminating them completely.
Finally,
Section~\ref{sec:Semantic Dependencies and Matching Up Stores}
demonstrates some challenges in determining which of a group of
stores is involved in a given semantic dependency.

@@

\subsubsection{Semantic Dependencies Can Be Many To One}
\label{sec:Semantic Dependencies Can Be Many To One}

@@

The exact definition of a computer language is subject to some debate,
with standards, implementations, and users all having some degree of
influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
each of which is subject to change over time.
In areas that are not well settled or where users might reasonably
want to resist the dictates of the standard,
compilers often provide switches to override their default behaviors.
An example is GCC's \co{-funsigned-char} command-line argument,
which causes it to treat variables of type \co{char} as unsigned.
More examples of user control over language semantics are given in
Appendix~\ref{app:User Influence Over Language Semantics}
and by the discussion
in~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.

@@

A noteworthy aspect of this transformation is that it invents an atomic load:
The original form of the code reads \co{x} only once,
whereas the transformed code reads it twice.
Therefore we can rule out transformations like this one by insisting
the compiler not invent atomic loads.
In fact, we will require that atomic accesses be treated as
``quasi volatile'', in that the compiler is allowed to omit,
merge, or reorder them but not invent them.

@@

We will say that one instruction is ordered after another to mean that
the CPU forces it to commit after the other one commits.
Hence instructions are ordered after the instructions they depend on.
(Note that this implies nothing about when a load instruction retrieves
its value from memory; it may do so long before it or a prior
instruction commits.)

@@

With the contents of the last few sections under our belts, we can
formulate the Fundamental Property that we would like all semantic
dependencies to satisfy.
Note that this formulation makes sense only for implementations in
which all atomic objects are treated as volatile or quasi volatile.
\begin{quote}
Let $W$ be a store which semantically depends on loads
$\{R_0$, \ldots,~$R_n\}$ in some abstract execution $A$,
and suppose that $W$ is not
omitted in some hardware execution $H$ realizing $A$.
Then for some $i$, load $R_i$ is not omitted in $H$ and the
instruction it maps to is ordered before the instruction $W$ maps to.
\end{quote}
It's quite straightforward to show that for any implementation in which
semantic dependencies satisfy the Fundamental Property, no OOTA cycle
has a nontrivial realization.

@@

If all the stores in the OOTA cycle are omitted then all the reads
must be omitted as well.
In effect, the entire cycle has been optimized
out of existence by the compiler.
Although we are unable to prove it, we conjecture that in this
situation there must be another abstract execution which has the
same observable effects as $A$ and is also realized by $H$, but in
which the OOTA cycle does not occur.
Thus there would be no way to tell, merely by observing the effects of
$H$, whether there was an OOTA cycle or not.
For this reason we declare realizations of OOTA cycles in which
all the accesses are omitted to be trivial.

@@

Computer systems are expected to continue increasing their use of
speculative execution.
Nevertheless, the main points of this paper will remain unaffected.
To see why, keep in mind that on real systems observable effects
must eventually be committed, but no effect can be committed
while any of the computations that determine the effect remain
speculative.
Combining this with the obvious fact that a machine instruction cannot
commit until it is no longer speculative, we can draw two conclusions:
\begin{itemize}
\item	A load cannot commit until the store it reads from has
	committed.\footnote{
	An early ARMv8 memory model did allow loads that read from an
	earlier store in the same thread to commit before the store.
	This apparent exception merely reflects a difference in
	nomenclature; in the memory model a store was said to commit at
	the time when it makes its new value available to other
	threads, whereas we say that a store commits at the time when
	the CPU has irrevocably decided that it will take place with a
	particular fully determined value and address.}
\item	A store cannot commit until all the loads on which it has a
	hardware dependency have committed.
\end{itemize}

@@

This paper has presented samples of code containing non-trivial
semantic dependencies, uncovering some shortcomings in typical definitions of
``semantic dependency''.
It pointed out examples where semantic dependencies
are a function of an execution rather than of the source code, and examples
where a single semantic dependency extends from multiple loads
to a store but not from any one of those loads.
These complex dependencies motivated a generalization of the definition
of ``OOTA cycle'', which is shown in
Section~\ref{sec:Semantic Dependencies Can Be Many To One}.

@@

\bibliographystyle{plain}
\bibliography{bib/RCU,bib/WFS,bib/hw,bib/os,bib/parallelsys,bib/patterns,bib/perfmeas,bib/refs,bib/syncrefs,bib/search,bib/swtools,bib/realtime,bib/TM,bib/standards,bib/memorymodel.bib}

>>>>

This section uncovers some semantic-dependency complexities.
Section~\ref{sec:Semantic Dependencies and Source Code}
shows that semantic dependencies are functions of executions,
rather than being strict functions of the source code.
Section~\ref{sec:Semantic Dependencies Can Be Many-To-One}
shows that semantic dependencies do not necessarily extend
from a single load to a single store, but can instead involve
multiple loads.
Section~\ref{sec:Semantic Dependencies Affected by Cross-Thread Optimizations}
shows that semantic dependencies can be affected by cross-thread
optimizations.
Sections~\ref{sec:Semantic Dependencies Affected by if Statements}
and~\ref{sec:Semantic Dependencies Not Affected by if Statements}
show that \co{if} statements can have surprising effects on semantic
dependencies, up to and including eliminating them completely.
Finally,
Section~\ref{sec:Semantic Dependencies and Matching Up Stores}
demonstrates some challenges in determining which of a group of
stores is involved in a given semantic dependency.

@@

\subsubsection{Semantic Dependencies Can Be Many-To-One}
\label{sec:Semantic Dependencies Can Be Many-To-One}

@@

The exact definition of a computer language is a subject of some debate,
with standards, implementations, and users all having some degree of
influence~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2},
and each being prone to change over time.
In areas that are not well settled or where users might reasonably
want to resist the dictates of the standard,
compilers often provide switches to override their default behaviors.
An example is GCC's \co{-funsigned-char} command-line argument,
which causes it to treat variables of type \co{char} as unsigned.
More examples of user control over language semantics are given in
Appendix~\ref{app:User Influence Over Language Semantics}
and by the discussion
in~\cite{KayvanMemarian2016DepthOfC-1,KayvanMemarian2016DepthOfC-2}.

@@

A noteworthy aspect of this transformation is that it invents an atomic load:
The original form of the code reads \co{x} only once,
whereas the transformed code reads it twice.
Therefore we can rule out transformations like this one by insisting
the compiler not invent (or duplicate) atomic loads.
In fact, we will require that atomic accesses be treated as
``quasi volatile'', in that the compiler is allowed to omit,
merge, or reorder them but not invent them.

@@

We will say that one instruction is \emph{ordered after} another to mean that
the CPU forces it to commit after the other one commits.
Hence instructions are ordered after the instructions they depend on.
(Note that this implies nothing about when a load instruction retrieves
its value from memory; it may do so long before it or a prior
instruction commits.)

@@

With the contents of the last few sections under our belts, we can
formulate the Fundamental Property that we would like all semantic
dependencies to satisfy.
Note that this formulation makes sense only for implementations in
which all atomic objects are treated as volatile or quasi volatile.
\begin{quote}
Let $W$ be a store which semantically depends on loads
$\{R_0$, \ldots,~$R_n\}$ in some abstract execution $A$,
and suppose that $W$ is not
omitted in some hardware execution $H$ realizing $A$.
Then for some $i$, load $R_i$ is not omitted in $H$ and the
instruction it maps to is ordered before the instruction $W$ maps to.
\end{quote}
It's quite straightforward to show that under any implementation in which
semantic dependencies satisfy the Fundamental Property, no OOTA cycle
has a nontrivial realization.

@@

If all the stores in the OOTA cycle are omitted then all the reads
must be omitted as well.
In effect, the entire cycle has been optimized
out of existence by the compiler.
Although we are unable to prove it, we conjecture that in this
situation there must be another abstract execution which has the
same observable effects as $A$ and is also realized by $H$, but in
which the OOTA cycle does not occur.
Thus there would be no way to tell, merely by observing the effects of
$H$, whether there was an OOTA cycle or not.
For this reason we declare realizations of OOTA cycles in which
all the accesses are omitted to be \emph{trivial}.

@@

Computer systems are expected to continue increasing their use of
speculative execution.
Nevertheless, the main points of this paper will remain unaffected.
To see why, keep in mind that on real systems observable effects
must eventually be committed, but no effect can be committed
while any of the computations that determine the effect remain
speculative.
Combining this with the obvious fact that a machine instruction cannot
commit until it is no longer speculative, we can draw two conclusions:
\begin{itemize}
\item	A load cannot commit until the store it reads from has
	committed.\footnote{
	An early ARMv8 memory model did allow loads that read from an
	earlier store in the same thread to commit before the store.
	This apparent exception merely reflects a difference in
	nomenclature; in the memory model a store was said to commit at
	the time when it made its new value available to other
	threads, whereas we say that a store commits at the time when
	the CPU has irrevocably decided that it will take place with a
	particular fully determined value and address.}
\item	A store cannot commit until all the loads on which it has a
	hardware dependency have committed.
\end{itemize}

@@

This paper has presented samples of code containing non-trivial
semantic dependencies, uncovering some shortcomings in typical definitions of
``semantic dependency''.
It pointed out examples where semantic dependencies
are a function of an execution rather than of the source code, and examples
where a single semantic dependency extends from multiple loads
to a store but not from any one of those loads.
These complex dependencies motivated a generalization of the definition
of ``OOTA cycle'', which is shown in
Section~\ref{sec:Semantic Dependencies Can Be Many-To-One}.

@@

\bibliographystyle{plain}
\addcontentsline{toc}{section}{References}
\bibliography{bib/RCU,bib/WFS,bib/hw,bib/os,bib/parallelsys,bib/patterns,bib/perfmeas,bib/refs,bib/syncrefs,bib/search,bib/swtools,bib/realtime,bib/TM,bib/standards,bib/memorymodel.bib}
